---
title: "Estill: Gene Methylation Enrichment Analysis"
author: "IVF meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

##  Introduction

Previously, Mandhri and I analysed the B-PROOF 450K data, trying to understand
whether vitamin supplementation caused changes in gene methylation.
We used Limma and some basic analyses, which showed no specific probes with FDR<0.05,
nor any DMRs.

In this analysis we will use the principle of Gene Set Enrichment Analysis, applying it
to many probes belonging to genes.
If the probes are trending in concert, then we can make some judgement about the
enrichment of those probes.
The statistical test used is the Wilcox test, which can be applied to competitive 
or self contained test types.

```{r,libs}

library("parallel")
library("dplyr")
library("kableExtra")
library("eulerr")
library("mitch")
library("tictoc")
source("meth_functions.R")

CORES= detectCores()/2

```

## Get array annotation data

```{r,anno}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

## Load Estill data

```{r,load}

nat_vs_fh <- readRDS("estill_nat_vs_fh.rds")
fh_vs_fz <- readRDS("estill_fh_vs_fz.rds")
nat_vs_fx <- readRDS("estill_nat_vs_fx.rds")
nat_vs_fz <- readRDS("estill_nat_vs_fz.rds")
iui_vs_fx <- readRDS("estill_iui_vs_fx.rds")
nat_vs_iui <- readRDS("estill_nat_vs_iui.rds")

if ( ! file.exists("ReactomePathways.gmt") ) {
  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip",
    destfile="ReactomePathways.gmt.zip")
  unzip("ReactomePathways.gmt.zip")
}
file.info("ReactomePathways.gmt")
genesets <- gmt_import("ReactomePathways.gmt")
length(genesets)

```

## Nat vs fh

```{r,nat_vs_fh}

dm <- nat_vs_fh$dma
head(dm,50) %>% kbl(caption = "Top significant genes with limma") %>% kable_paper("hover", full_width = F)
dm <- dm[,c("UCSC_RefGene_Name","t")]
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc() 
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res,file="estill_gmeawg_nat_vs_fh.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")
head(mres$enrichment_result,20) %>% kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeawg_mitch_nat_vs_fh.html",overwrite=FALSE)
# Promoter
dm <- nat_vs_fh$dma
dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]
dm <- dm[,c("UCSC_RefGene_Name","t")]
head(dm,50) %>% kbl(caption = "Top significant promoters with limma") %>% kable_paper("hover", full_width = F)
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]] 
write.table(res ,file="estill_gmeapr_nat_vs_fh.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")
head(mres$enrichment_result,20) %>%  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeapr_mitch_nat_vs_fh.html",overwrite=FALSE)
rm(nat_vs_fh)

```

## fh vs fz

```{r,fh_vs_fz}

dm <- fh_vs_fz$dma
head(dm,50) %>% kbl(caption = "Top significant genes with limma") %>% kable_paper("hover", full_width = F)
dm <- dm[,c("UCSC_RefGene_Name","t")]
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res,file="estill_gmeawg_fh_vs_fz.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")
head(mres$enrichment_result,20) %>% kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeawg_mitch_fh_vs_fz.html",overwrite=FALSE)
# Promoter
dm <- fh_vs_fz$dma
dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]
dm <- dm[,c("UCSC_RefGene_Name","t")]
head(dm,50) %>% kbl(caption = "Top significant promoters with limma") %>% kable_paper("hover", full_width = F)
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res ,file="estill_gmeapr_fh_vs_fz.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")
head(mres$enrichment_result,20) %>%  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeapr_mitch_fh_vs_fz.html",overwrite=FALSE)
rm(fh_vs_fz)

```

## nat vs fx

```{r,nat_vs_fx}

dm <- nat_vs_fx$dma
head(dm,50) %>% kbl(caption = "Top significant genes with limma") %>% kable_paper("hover", full_width = F)
dm <- dm[,c("UCSC_RefGene_Name","t")]
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res,file="estill_gmeawg_nat_vs_fx.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")
head(mres$enrichment_result,20) %>% kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeawg_mitch_nat_vs_fx.html",overwrite=FALSE)
# Promoter
dm <- nat_vs_fx$dma
dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]
dm <- dm[,c("UCSC_RefGene_Name","t")]
head(dm,50) %>% kbl(caption = "Top significant promoters with limma") %>% kable_paper("hover", full_width = F)
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res ,file="estill_gmeapr_nat_vs_fx.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")
head(mres$enrichment_result,20) %>%  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeapr_mitch_nat_vs_fx.html",overwrite=FALSE)
rm(nat_vs_fx)

```

## nat vs fz

```{r,nat_vs_fz}

dm <- nat_vs_fz$dma
head(dm,50) %>% kbl(caption = "Top significant genes with limma") %>% kable_paper("hover", full_width = F)
dm <- dm[,c("UCSC_RefGene_Name","t")]
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res,file="estill_gmeawg_nat_vs_fz.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")
head(mres$enrichment_result,20) %>% kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeawg_mitch_nat_vs_fz.html",overwrite=FALSE)
# Promoter
dm <- nat_vs_fz$dma
dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]
dm <- dm[,c("UCSC_RefGene_Name","t")]
head(dm,50) %>% kbl(caption = "Top significant promoters with limma") %>% kable_paper("hover", full_width = F)
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res ,file="estill_gmeapr_nat_vs_fz.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")
head(mres$enrichment_result,20) %>%  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeapr_mitch_nat_vs_fz.html",overwrite=FALSE)
rm(nat_vs_fz)

```

## nat_vs_iui

```{r,nat_vs_iui}

dm <- nat_vs_iui$dma
head(dm,50) %>% kbl(caption = "Top significant genes with limma") %>% kable_paper("hover", full_width = F)
dm <- dm[,c("UCSC_RefGene_Name","t")]
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res,file="estill_gmeawg_nat_vs_iui.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")
head(mres$enrichment_result,20) %>% kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeawg_mitch_nat_vs_GIFT.html",overwrite=FALSE)
# Promoter
dm <- nat_vs_iui$dma
dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]
dm <- dm[,c("UCSC_RefGene_Name","t")]
head(dm,50) %>% kbl(caption = "Top significant promoters with limma") %>% kable_paper("hover", full_width = F)
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res ,file="estill_gmeapr_nat_vs_iui.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")
head(mres$enrichment_result,20) %>%  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeapr_mitch_nat_vs_GIFT.html",overwrite=FALSE)
rm(nat_vs_iui)

```


## iui_vs_fx

```{r,iui_vs_fx}

dm <- iui_vs_fx$dma
head(dm,50) %>% kbl(caption = "Top significant genes with limma") %>% kable_paper("hover", full_width = F)
dm <- dm[,c("UCSC_RefGene_Name","t")]
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res,file="estill_gmeawg_iui_vs_fx.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")
head(mres$enrichment_result,20) %>% kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeawg_mitch_nat_vs_GIFT.html",overwrite=FALSE)
# Promoter
dm <- iui_vs_fx$dma
dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]
dm <- dm[,c("UCSC_RefGene_Name","t")]
head(dm,50) %>% kbl(caption = "Top significant promoters with limma") %>% kable_paper("hover", full_width = F)
hist(dm$t)
tic() ; gmea <- calc_sc(dm) ; time2 <- toc()
df <- gmea[[1]]
res <- gmea[[2]]
write.table(res ,file="estill_gmeapr_iui_vs_fx.tsv")
head(res,50) %>% kbl(caption = "Top significant genes with GMEA") %>% kable_paper("hover", full_width = F)
gmea_volc(res)
gmea_barplot(res)
dmscore <- data.frame( res$median * res$sig)
rownames(dmscore) <- rownames(res)
colnames(dmscore) <- "metric"
mres <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")
head(mres$enrichment_result,20) %>%  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>% kable_paper("hover", full_width = F)
#mitch_report(mres,outfile="gmeapr_mitch_nat_vs_GIFT.html",overwrite=FALSE)
rm(iui_vs_fx)

```

## Session Information

For reproducibility.

```{r,sessioninfo}

sessionInfo()

```
